\providecommand{\main}{..}  
% *Modification: redefine path location, must go before \documentclass
\documentclass[../computer-history.tex]{subfiles}

\begin{document}
From here is where the conflict with INTEL begins especially with the latter’s creation of a microprocessor in 1974. AMD then followed in 1975 after reverse engineering the INTEL 8080 (Intel’s second microprocessor) the AM9080. In 1978 when INTEL entered into a contract with IBM who, due to the sheer volume of processors required, requested that this contract would be completed through the use of contractors, with AMD among those chosen. This secured AMD’s entry into the x86 processor market and created a technology exchange agreement with Intel so the smaller company could produce the product at IBM’s desired specification. Thus AMD began production of the AM286, a copy of the INTEL80826, which held the advantage of much higher clock speeds (Intel resting at around 12.5 MHz and AMD at 20 MHz). This, in addition to the 1975 AM9080, would start a trend between AMD and Intel where Intel would produce a device and AMD, through their agreement, would take the design and create a clone which would have a much higher clock speed. The next notable release was the AM486 and the AM5x86, released in 1994 and 1985 respectively. Both devices ran the same architecture but the AM5x86 ran much faster at 133 MHz (or potentially 150 MHz) as compared to the AM486’s speed of 120 MHz. These devices were the final Intel clones and marked the beginning of AMD creating CPUs of their own design and a separation from Intel. %........ Emma's Research here.....% 
\cite{Ryzen}Amd ryzen was an evolutionary step in major architecture since Bulldozer. Ryzen, codenamed Zen offered 1331 pin AM4 socket type for motherboards which supported DDR4 RAM, PCIe Gen 3, enhanced sound, SATA express, and USB 3.1 gen2 10gbs. Ryzen CPU's were the first chips in AMD to be developed at 14nm, and offered integrated graphics which were mainly used for laptops.\cite{Ryzen2} AMD has developed 7 generations of CPUs since 2016 with generation 4, 5, and 6 being for laptops only, AMD also developed 4 models, Ryzen 3, 5, 7, and 9 which stayed the same but had an increase in performance for each generation. Each model does a different task with Ryzen 3 being general work such as PowerPoint and net browsing and Ryzen 9 being used for graphic design, high game performance, and other heavy tasks, Ryzen 5 and 7 were meant to be best for gaming and multitasking.\cite{Ryzen3} Ryzen series beat AMDs old CPU technology FX series by 50 percent with its Zen 3 technology.\cite{Ryzen4} Zen 3 offered a complete redesign with 20 major changes such as wider and more flexible execution resources such as, load and store bandwidth to help feed execution, and 
a streamlined front-end to get more threads in flight; Higher Instruction Per Clock which could extract an average of 19 percent additional performance for every MHz of frequency; Lower Latency with 8 cores and 32MB of L3 cache which reduces core-to-core and core-to-cache latencies, which rivaled Zen 2 cache by two times as much; And Incredible efficiency, new Zen 3 architecture such as AMD Ryzen 5000 series gave a 24 percent generational improvement in energy efficiency 
and a 2.8 times lead over competing architectures making this CPU one of the top CPUs in the world.

\onecolumn
\\\textbf{1982 INTEL 8086 clone}\\ \\
\begin{tabular}{c|c}
    Release Date & 1982 \\
    Architecture & 16-bit \\
    Data Bus & 16-bit \\
    Address Bus & 20-bit 
    Maximum Memory Support & 1 MB \\
    L1 Cache & None \\
    L2 Cache & None \\
    Frequency & 4 - 10 MHz \\
    FSB & 4 - 10 MHz \\
    FPU & 8087 (Sold Separate) \\
    SIMD & None  \\
    Fab & 3000 nm \\
    Transistor Count & 29,000 \\
    Power Consumption & N/A \\
    Voltage & 5 V \\
    Die Area & 33 mm² \\
    Socket &40 pins
\end{tabular}

\ \\ \\
\textbf{AMD Bulldozer Zambezi} \\ \\
\begin{tabular}{c|c}
    Code Name & Zambezi \\
    Date & October 2011 \\
    Architecture & 64-bit \\
    Data Bus & 64-bit \\
    Address Bus & 64-bit \\
    Maximum Memory Support & 1 TB \\
    L1 Cache & 64 KB + (2 x 16 KB) \\
    L2 Cache & 2 MB (Full Speed) \\
    L3 Cache & 8 MB \\
    Clock Speed & 2.8 - 4.2 GHz (4.3 GHz Turbo) \\
    Memory Controller & Dual-Channel DDR3-1866 \\
    HyperTransport & 2600 MHz \\
    Core Count & 4, 6, 8 \\
    SIMD & MMX, SSE, SSE2, SSE3, SSSE3, SSE4a, SSE4.1/4.2, AVX \\
    Instructions & AES, FMA4, XOP \\
    Fab & 32 nm \\
    Transistor Count & N/A \\
    Power Consumption & 95 - 125 W \\
    Voltage & 0.95 - 1.4125 V \\
    Die Area & 316 mm² \\
    Socket & AM3+
\end{tabular}

\ \\ \\
\textbf{AMD Ryzen 9 7950X} \\ \\
\begin{tabular}{c|c}
    Core name & Raphael \\
    Release date & September 27, 2022 \\
    Generation & 5th \\
    Architecture & Zen 4 \\
    Frequency & 4.5 GHz \\
    L1 Cache & 1MB \\
    L2 Cache & 16MB \\
    L3 Cache & 64MB \\
    Cores & 16 \\
    Threads & 32 \\
    Unlocked & Yes \\
    Socket & AM5 \\
    Thermal Solution & Not included \\
    Thermal Design Power & (TDP)	170 W \\
    Max. Temperature & 95 °C \\
    Word Size & 64 bit \\
    TSMC FinFET process (Lithography) & 5 nm \\
    Transistor count & 12600 million \\
\end{tabular}
\two coulmn
%********************************************************************************
%      Bibliography
%********************************************************************************
\biblio
\end{document}
